---
title: "Public Health and Economic Impact of Severe Weather Events in the U.S."
author: "Amade Annissimo"
date: "November 18, 2015"
output: html_document
---

## Synopsis

## Data processing

Dataset for the analysis comes from the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.

```{r, echo=TRUE, cache = TRUE}

stormDataFile <- "weather.rds"

if(!file.exists(stormDataFile)) {
    fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    destFile <- "weather.csv.bz2"
    download.file(fileUrl, destfile = destFile, method = "curl")
    storm <- read.csv(destFile)
    saveRDS(storm,stormDataFile)
    }
storm <- readRDS(stormDataFile)
```

The data set consist of `r ncol(storm)` variables and `r format(nrow(storm), big.mark=",", scientific=FALSE)` observations.

Data processing is initially performed on a random sample of 1000 observation for faster processing. At the later stage, preprocessing methods will be applied to the entire dataset.

```{r, echo=TRUE, cache = TRUE}
set.seed(1000)
stormSample <- storm[sample(nrow(storm), 1000), ]
```

Use the 'purrr' package to convert the factor to charaters for easier processing.


```{r, echo=TRUE}
library(purrr)
stormSample %>% map_if(is.factor, as.character) -> stormSample
```

Store remove 'remarks' variable from the main dataset for easier processing and store it to a separete variable

```{r, echo=TRUE}
remarks <- stormSample$REMARKS
library(dplyr, warn.conflicts = FALSE) # supress the messages on conflicst in variables
stormSample <- select(stormSample, -REMARKS)
```
Remove other variables which won't be used in the analysis
```{r, echo=TRUE}
#WFO - weather forecast office
stormSample <- select(stormSample, -c(ZONENAMES, BGN_TIME, END_TIME, TIME_ZONE,
                                      STATE__, COUNTY, COUNTYNAME, BGN_RANGE,
                                      BGN_AZI, BGN_LOCATI, COUNTY_END, COUNTYENDN,
                                      END_RANGE, END_AZI, END_LOCATI, LENGTH,
                                      WIDTH, WFO))
```

Dates were processed using the 'lubridate' package.

```{r, echo=TRUE}
library(lubridate)
stormSample <- stormSample %>% mutate(
    BGN_DATE = as.Date(mdy_hms(stormSample$BGN_DATE)),
    END_DATE = as.Date(mdy_hms(stormSample$END_DATE))
    )
```

Chect the number of events (all types), by year, to see if a cutoff at 1996 is possible.

## Results